Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance

### 2013-12-16 HRUIofMovingObjects/jrnl_article.tex
Moving objects: done some writing

\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}. The method has been shown to improve the lateral resolution and edge sharpness \cite{Synnevag2007, Synnevag2009, Chen2011}. This is accomplished by adaptively adjusting the array weights based on element-to-element covariance in order to minimize interference. The method is, however, also highly sensitive to model errors \cite{Mestre2006, Widrow1982, Wax1996, Wax1996a}. This has led to the development of several robust versions of the Capon beamformer with e.g. spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying a steering vector uncertainty set \cite{Lorenz2005, Rubsamen2013}. Spatial smoothing and diagonal loading, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}. The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution parameters \cite{Synnevag2007a}. Steering vector uncertainty sets, however, have not been applied for medical ultrasound imaging.

@@

In \cite{Cox1973}, Cox presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations. He also states the following important fact about the Capon beamformer [the $\vec{k}_3$ processor]: 
\begin{quote}
"\textit{... a $\vec{k}_3$ processor requires more closely spaced beams than a ... $\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}", 
\end{quote}
where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer. In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}. A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to two times the required beam spacing using narrow band excitation. Located in the farfield of the sector is an object moving parallel to the array surface. The plot shows the scanned output when the object is located at plus and minus half the required beam spacing for DAS [See (\ref{eq:resolution})]. %No spatially white noise is present. Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity. 
The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output. A diagonal loading of $1/100$ corresponds to a signal-to-noise ratio (SNR) of 20 dB before array gain is applied. The figure shows that a DAS beamformer, with the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array. However if the Capon beamformer is used with the same beam density the loss could be as large as $27$ dB. This imaging system is then highly sensitive to small spatial shifts. A $1.9$ dB scalloping loss for the Capon beamforming is obtained for this specific example when the beam spacing or object distance is decreased with a factor of 25.

@@

In this paper we investigate the issue further. We show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS, the predicted scalloping loss is observed both in simulations and \textit{in-vitro} data. The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate and processing throughout. We try two answer this by studying the scalloping loss of the Capon beamforming and investigating different methods aimed at reducing it. We also discuss how the different methods will affect the acquisition frame rate and the processing complexity.

In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and shift-invariance. In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before different methods for improved local lateral shift-invariance are presented in Section \ref{sec:methods}. Then, in Section \ref{sec:res} we present the results of applying the most promising method on \textit{in-vitro} data. Finally we discuss our findings in Section \ref{sec:dis} and give our conclusions in Section \ref{sec:con}.

@@

Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}). The dotted vertical lines indicates the position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway. The probe was excited with a 2.5 MHz and 1.5 cycles long pulse. The transmit focus in azimuth and elevation was configured to be equal and the point scatterer was moved through this focus. We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE). We therefore conclude that the system is locally laterally shift-invariant. %In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.

@@

\subsection{Oversampling by Phase Rotation}
The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation. The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since the ultrasound data are pre-delayed. However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors. In a narrow band setting the beam can make a sweep across the whole imaging sector by using phase rotation. This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius}). The maximal steering angle is therefore given by
\begin{align}
\theta_{max} = \alpha p \Delta,
\end{align}
where $\alpha$ is the pulse length in $\lambda$'s, and $p$ and $\Delta$ is the oversampling factor and beam spacing respectively as defined in (\ref{eq:resolution}). For our simulations $\theta_{max} = 3\Delta$.
It is therefore possible to make a sweep using phase rotation between a given beam and half-way to its two nearest neighbours without introducing large errors. Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.

@@

\section{Discussion}\label{sec:dis}
In this paper we have investigated the lateral shift-invariance property of the Capon beamformer when imaging moving objects. We have shown that the Capon beamformer is less sensitive to small lateral shifts if oversampling based on phase rotation is applied. This without reducing the acquisition frame rate and with a minor increase in processing complexity. %In Section \ref{sec:capon_LLSI} it was shown that a Capon beamformer with high resolution parameters is not locally laterally shift-invariant when subjected to motion. The root cause was found to be too coarse lateral sampling. 
%In Section \ref{sec:methods} we presented methods for improved shift-invariance based on oversampling. The most promising method, with respect to improved shift-invariance, reduction in acquisition frame rate and processing complexity combined, was based on phase rotation. 
%In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data when the ultrasound probe is subjected to motion. 
In simulations the method reduced the maximal scalloping loss by more than 20 dB. On \textit{in-vitro} data the reduction was more than 7 dB.

Oversampling is not only essential to keep the amplitude level of moving point scatterers. It should also helps to improve contrast and edge definitions. If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of a moving object will fluctuate in the same way as a moving point scatterer would have sparkled.

@@

Faced with an imaging scene containing bright point scatterers or sharp edges, one could wonder how large the oversampling factor needs to be in order to avoid visible scalloping losses. In his seminal paper, Cox \cite{Cox1973} derives the required beam density for a given output SNR. For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}). Since resolution depends on SNR we would therefore recommend to first use the dynamic range in the image as guidance. The amount of diagonal loading will also dictate the maximal achievable resolution since $10\log{d}$ can be interpreted as the input SNR seen by the Capon beamformer. %Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves. For ultrasound data only a 50 \% increase in resolution was found in \cite[Fig.~7]{Asen} for a 30 dB output SNR when measured in a similar way as Cox [6 dB saddle point]. However, sixteen times oversampling was required to make the imaging system shift invariant in the same paper. In this paper more than ten times oversampling was found to be required for a 40 dB output SNR (after array gain) in simulations, however for the \textit{in-vitro} data ten times oversampling was enough for the system to be locally laterally shift-invariant. This might be a result of improved initial sampling (as discussed in Section \ref{sec:res}) and a higher initial noise level before diagonal loading was applied.

@@

Oversampling by phase rotation can therefore also introduce similar block artifacts as parallel receive beamforming (PRB). These artifacts should be possible to remove in the same way as for PRB by using the synthetic transmit beamforming (STB) approach. If we distribute our steering angles in an overlapping scheme [like rx-beams are spread out for STB] we get a phase rotation version of STB where e.g. eight phase rotated beams are averaged into four. The geometric distortion, or stripes, seen at the bottom of the sector in the video \multimedia{Capon + 10PR} should be possible to remove using this technique.

>>>>

\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}. The method has been shown to improve the lateral resolution and edge sharpness \cite{Synnevag2007, Synnevag2009, Chen2011}. This is accomplished by adaptively adjusting the array weights based on element-to-element covariance in order to minimize interference. The method is, however, also highly sensitive to model errors \cite{Mestre2006, Widrow1982, Wax1996, Wax1996a}. This has led to the development of several robust versions of the Capon beamformer with e.g. spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying steering vector uncertainty sets \cite{Lorenz2005, Rubsamen2013}. Spatial smoothing and diagonal loading, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}. The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution parameters \cite{Synnevag2007a}. Steering vector uncertainty sets, however, have not been applied for medical ultrasound imaging.

@@

In \cite{Cox1973}, Cox presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations. He also states the following important fact about the Capon beamformer [the $\vec{k}_3$ processor]: 
\begin{quote}
"\textit{... a $\vec{k}_3$ processor requires more closely spaced beams than a ... $\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}", 
\end{quote}
where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer. In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}. A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to four times the array resolution using narrow band excitation. Located in the farfield of the sector is an object moving parallel to the array surface. The plot shows the high precision scanned output when the object is located at plus and minus half the required beam spacing for DAS [See (\ref{eq:resolution})]. %No spatially white noise is present. Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity. 
The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has been added to the Capon beamformer output. A diagonal loading of $1/100$ corresponds to a signal-to-noise ratio (SNR) of 20 dB before array gain is applied. The figure shows that a DAS beamformer, using the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array. However if the Capon beamformer is used with the same beam density the loss could be as large as $27$ dB. This imaging system is then highly sensitive to small spatial shifts, either by the object or the array itself. A $1.9$ dB scalloping loss for the Capon beamforming is obtained for this specific example when the beam spacing or object distance is decreased with a factor of 25.

@@

In this paper we investigate the issue further. We show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS, the predicted scalloping loss is observed both in simulations and \textit{in-vitro} data. The question is then, what can we do in order to improve on the situation without affecting the acquisition frame rate and processing throughout. %We try two answer this by studying the scalloping loss of the Capon beamforming and investigating different methods aimed at reducing it. We also discuss how the different methods will affect the acquisition frame rate and the processing complexity.

The next section gives an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and shift-invariance. In Section \ref{sec:capon_LLSI} we study the scalloping loss yield by the Capon beamformer and its local lateral shift-invariance property.  Different methods for reduced scalloping loss and by that improved local lateral shift-invariance are then presented in Section \ref{sec:methods}. Then, in Section \ref{sec:res} we present the results of applying the most promising method on \textit{in-vitro} data. Finally we discuss our findings in Section \ref{sec:dis} and give our conclusions in Section \ref{sec:con}.

@@

Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}). The dotted vertical lines indicates the position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway. The probe was excited with a 2.5 MHz and 1.5 cycles long pulse. The transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point. We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE). We therefore conclude that the system is locally laterally shift-invariant. %In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.

@@

\subsection{Oversampling by Phase Rotation}
The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation. The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since ultrasound data have to be pre-delayed. However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors. In a narrow band setting the beam can make a sweep across the whole imaging sector by phase steering. This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius}). The maximal steering angle is therefore given by
\begin{align}
\theta_{max} = \alpha p \Delta,
\end{align}
where $\alpha$ is the pulse length in $\lambda$'s, and $p$ and $\Delta$ is the oversampling factor and beam spacing respectively as defined in (\ref{eq:resolution}). For our simulations $\theta_{max} = 3\Delta$.
It is therefore possible to make a sweep using phase rotation between a given beam and half-way to its two nearest neighbours without introducing large errors. Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.

@@

\section{Discussion}\label{sec:dis}
In this paper we have investigated the lateral shift-invariance property of the Capon beamformer when imaging moving objects. We have shown that the Capon beamformer is less sensitive to small lateral shifts if oversampling based on phase rotation is applied. This without reducing the acquisition frame rate and with a minor increase in processing complexity. %In Section \ref{sec:capon_LLSI} it was shown that a Capon beamformer with high resolution parameters is not locally laterally shift-invariant when subjected to motion. The root cause was found to be too coarse lateral sampling. 
%In Section \ref{sec:methods} we presented methods for improved shift-invariance based on oversampling. The most promising method, with respect to improved shift-invariance, reduction in acquisition frame rate and processing complexity combined, was based on phase rotation. 
%In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data when the ultrasound probe is subjected to motion. 
In simulations the method reduced the maximal scalloping loss by more than 20 dB. On \textit{in-vitro} data the reduction was more than 7 dB. On \textit{in-vitro} data the imaging system is almost locally laterally shift invariant.

Note that oversampling is not only essential to keep the amplitude level of moving point scatterers. It should also helps to improve contrast and edge definitions. If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of a moving object will fluctuate in the same way as a moving point scatterer would have sparkled.

@@

Faced with an imaging scene containing bright point scatterers or sharp edges, one could wonder how large the oversampling factor needs to be in order to avoid visible scalloping losses. In his seminal paper, Cox \cite{Cox1973} derives the required beam density for a given output SNR. For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}). Since resolution for the Capon beamformer depends on SNR we would recommend to first use the dynamic range in the image as guidance. The amount of diagonal loading will also dictate the maximal achievable resolution since $10\log{d}$ can be interpreted as the maximal input SNR seen by the Capon beamformer. %Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves. For ultrasound data only a 50 \% increase in resolution was found in \cite[Fig.~7]{Asen} for a 30 dB output SNR when measured in a similar way as Cox [6 dB saddle point]. However, sixteen times oversampling was required to make the imaging system shift invariant in the same paper. In this paper more than ten times oversampling was found to be required for a 40 dB output SNR (after array gain) in simulations, however for the \textit{in-vitro} data ten times oversampling was enough for the system to be locally laterally shift-invariant. This might be a result of improved initial sampling (as discussed in Section \ref{sec:res}) and a higher initial noise level before diagonal loading was applied.

@@

Oversampling by phase rotation can therefore introduce similar block artifacts as parallel receive beamforming (PRB). These artifacts should be possible to remove in the same way as for PRB by using the synthetic transmit beamforming (STB) technique. If we distribute our steering angles in an overlapping scheme [like rx-beams are spread out for STB] we get a phase rotation version of STB where e.g. eight phase rotated beams are averaged into four. The geometric distortion, or stripes, seen at the bottom of the sector in the video \multimedia{Capon + 10PR} should be possible to remove using this technique.
